---
title: "Week 6 Homework"
author: "Yurun (Ellen) Ying"
date: '2022-06-24'
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(rethinking); library(dagitty)
```

## Course homework

### Problem 1

Prior predictive simulation for the tank survival probabilities:

$\mathrm{logit}(p_i) = \alpha_{TANK[i]}$

$\alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma)$

$\bar{\alpha} \sim \mathrm{Normal}(0, 1)$

$\sigma \sim \mathrm{Exponential}(1)$

```{r}
set.seed(7)
n <- 1e5
a_bar <- rnorm(n, 0, 1)
sigma <- rexp(n, 1)
a <- rnorm(n, a_bar, sigma)
dens(inv_logit(a), adj = 0.1, lwd = 2, xlab = "survival probability")
```

Write the above code as a function and plot for other prior values of sigma.

```{r, include = FALSE}
multi_prior <- function(lambda = 1){
  set.seed(7)
  n <- 1e5
  a_bar <- rnorm(n, 0, 1)
  sigma <- rexp(n, lambda)
  a <- rnorm(n, a_bar, sigma)
  dens(inv_logit(a), adj = 0.1, lwd = 2, xlab = "survival probability")
  mtext(paste0("sigma ~ exp(", lambda, ")"))
}
```

```{r}
par(mfrow = c(2, 2))
for (lambda in c(0.1, 1, 5, 10)) multi_prior(lambda)
```

Too tight a prior makes the model believe that the survival probability is either 0 or 1. Relaxing the prior makes the model assign most probability mass around the center before it is fed with the data.


### Problem 2

Start with the varying intercept model and add `pred` and `size` as predictors. Consider their interaction.

```{r}
# a DAG
dag1 <- dagitty("dag{Pred <- Size -> Prob; Pred -> Prob}")
coordinates(dag1) <- list(x = c(Pred = 0.2, Size = 0.5, Prob = 0.8),
                          y = c(Pred = 1, Size = 0.5, Prob = 1))
drawdag(dag1)
```

Size and predators are likely to impact survival probability independently. Size can also impact how much effect predators are present can influence the probability.

#### Correction

This is an experiment. Manipulations don't obviously confound each other. So instead we need to use this DAG.

```{r}
dag1.1 <- dagitty("dag{Size -> Prob; Pred -> Prob; Tank -> Prob; Dens -> Prob}")
coordinates(dag1.1) <- list(x = c(Pred = 0.2, Size = 0.3, Prob = 0.5, Dens = 0.8, Tank = 0.7),
                          y = c(Pred = 1, Size = 0.8, Prob = 1, Dens = 1, Tank = 0.8))
drawdag(dag1.1)
```



```{r}
data("reedfrogs")
d <- reedfrogs

# a clean data
d$tank <- 1:nrow(d)
dat <- list(
  tank = d$tank,
  N = d$density,
  surv = d$surv,
  S = ifelse(d$size == "small", 1, 2), # 1 = small, 2 = big
  P = ifelse(d$pred == "no", 1, 2) # 1 = no, 2 = pred
)
```

Fit a model:

$S_i \sim \mathrm{Binomial}(N_i, p_i)$

$\mathrm{logit}(p_i) = \alpha_{TANK[i]} + \beta_{[P,S]}$

$\alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma)$

$\bar{\alpha} \sim \mathrm{Normal}(0, 1.5)$

$\sigma \sim \mathrm{Exponential}(1)$

$\beta_{[P,S]} \sim \mathrm{Normal}(0, 1)$

```{r}
# priors
# set.seed(7)
# n <- 1e5
# a_bar <- rnorm(n, 0, 1)
# sigma <- rexp(n, 1)
# a <- rnorm(n, a_bar, sigma)
# b <- rnorm(n, 0, 1)
# p <- a + b
# dens(inv_logit(p), adj = 0.1, lwd = 2, xlab = "survival probability")

m2 <- ulam(
  alist(
    surv ~ dbinom(N, p),
    logit(p) <- a[tank] + b[P,S],
    a[tank] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1),
    sigma ~ dexp(1),
    matrix[P,S]:b ~ dnorm(0, 1)
  ), data = dat, chains = 4, cores = 4
)

precis(m2, 3, pars = c("a_bar", "sigma", "b"))
```

From the table, we can roughly see that predators being present decreases survival rate, and being large in size has different effects depending on whether predators is present or not.

```{r}
set.seed(509)
m2_post <- extract.samples(m2)
diff_pred_small <- inv_logit(m2_post$b[,2,1]) - inv_logit(m2_post$b[,1,1])
diff_pred_big <- inv_logit(m2_post$b[,2,2]) - inv_logit(m2_post$b[,1,2])
dens(diff_pred_small, col = 2, lwd = 3,
     xlab = "Predator P-A contrast", xlim = c(-0.8,-0.1), ylim = c(0,7))
dens(diff_pred_big, col = 4, lwd = 3, add = TRUE)
text(-0.3, 3, "Small size", col = 2)
text(-0.75, 3, "Big size", col = 4)
```

When the size is small, predators being present decreases the survival probability by around 0.4; when the size is big, predators being present decreases the survival probability by around 0.6.

```{r}
diff_size_np <- inv_logit(m2_post$b[,1,2]) - inv_logit(m2_post$b[,1,1])
diff_size_p <- inv_logit(m2_post$b[,2,2]) - inv_logit(m2_post$b[,2,1])
dens(diff_size_np, col = 2, lwd = 3,
     xlab = "Size B-S contrast", xlim = c(-0.5,0.4), ylim = c(0,7))
dens(diff_size_p, col = 4, lwd = 3, add = TRUE)
text(0.18, 3, "Predators\nabsent", col = 2)
text(-0.3, 3, "Predators\npresent", col = 4)
abline(v = 0, lty = 2)
```

Whether predators are absent, having big size has little effect on the survival probability; when predators are present, having big size decreases the survival probabilities by around 0.2.


### Problem 3

Estimate the causal effect of `density` on survival. Consider whether `pred` modifies the effect of density.

We will be using the following model:

$S_i \sim \mathrm{Binomial}(N_i, p_i)$

$\mathrm{logit}(p_i) = \alpha_{TANK[i]} + \beta_{[P,S]} + \beta_{D[P]}D$

$\alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma)$

$\bar{\alpha} \sim \mathrm{Normal}(0, 1.5)$

$\sigma \sim \mathrm{Exponential}(1)$

$\beta_{[P,S]} \sim \mathrm{Normal}(0, 1)$

$\beta_{D[P]} \sim \mathrm{Normal}(0, 1)$

```{r}
# a new dataset
dat2 <- list(
  tank = d$tank,
  N = d$density,
  surv = d$surv,
  S = ifelse(d$size == "small", 1, 2), # 1 = small, 2 = big
  P = ifelse(d$pred == "no", 1, 2), # 1 = no, 2 = pred
  D = standardize(d$density)
)

# fit a model
m3 <- ulam(
  alist(
    surv ~ dbinom(N, p),
    logit(p) <- a[tank] + bPS[P,S] + bD[P]*D,
    a[tank] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1),
    sigma ~ dexp(1),
    bD[P]~ dnorm(0, 1),
    matrix[P,S]:bPS ~ dnorm(0, 1)
  ), data = dat2, chains = 4, cores = 4
)

precis(m3, 2, pars = c("a_bar", "sigma", "bPS", "bD"))
```

The sigma value in this model has a lower mean value than the sigma in the model for last problem. This is probably because this model has one more predictor and explains more variance in the outcome through the predictors other than the intercepts. Therefore, the variance left unexplained for the intercepts is lower, resulting in a lower estimate of $\sigma$.


```{r}
set.seed(509)
m3_post <- extract.samples(m3)
dens(inv_logit(m3_post$bD[,1]), col = 2, lwd = 3,
     xlab = "BetaD", xlim = c(0.2, 0.8), ylim = c(0,10))
dens(inv_logit(m3_post$bD[,2]), col = 4, lwd = 3, add = TRUE)
text(0.68, 3, "Predators\nabsent", col = 2)
text(0.25, 3, "Predators\npresent", col = 4)
```

When predators are absent, increasing the population density by 10 can increase the survival probability by around 0.55; when predators are present, increasing the population density by 10 can only increase the survival probability by around 0.4.


### Problem 4

Compute the expected causal effect of removing predators.

We need to average over all the predictors. We will assign equal number of cases to the four conditions. Each condition will contain tank number 1-48.

```{r}
# repeat all the tank number four times
tank <-  rep(1:nrow(d), 4)
# half have 10 and the other have 35 individuals
N <- c(rep(10, nrow(d)*2), rep(35, nrow(d)*2))
# standardize N using the mean and sd of the empirical data
D <- (N - attr(dat2$D,"scaled:center"))/attr(dat2$D,"scaled:scale")
# half small and half large
S <- rep(c(rep(1, nrow(d)), rep(2, nrow(d))), 2)

# simulate as if predators are missing for all tanks
sim_preda <- 
  link(m3, data = data.frame(tank = tank, N = N, D = D, S = S, P = rep(1, nrow(d)*4)))
# simulate as if predators are present for all tanks
sim_predp <- 
  link(m3, data = data.frame(tank = tank, N = N, D = D, S = S, P = rep(2, nrow(d)*4)))
diff_ap <- inv_logit(sim_preda) - inv_logit(sim_predp)
dens(diff_ap, lwd = 2, xlab = "Counterfactual predators A-P contrast")
```

Removing predators will increase the average survival rate across all conditions by about `r round(mean(diff_ap), 2)``.


#### Correction

What we would like to do for the varying intercepts is instead to randomly sample from the posterior distribution and include them in the linear model.

```{r}
n <- nrow(d)
N <- rep(c(10, 10, 35, 35), times = n)
D <- (N - attr(dat2$D,"scaled:center"))/attr(dat2$D,"scaled:scale")
S <- rep(c(1, 2, 1, 2), times = n)

# extract sample of the varying intercept
post <- extract.samples(m3)
aT <- replicate(length(D), rnorm(2e3, post$a_bar, post$sigma))

# simulate as if predators are missing for all tanks
P <-  rep(rep(1, 4), times = n)
sim_preda <- sapply(
  1:length(D),
  function (i) inv_logit(aT[,i] + post$bPS[,P[i],S[i]] + post$bD[,P[i]] * D[i]))
# simulate as if predators are present for all tanks
P <-  rep(rep(2, 4), times = n)
sim_predp <- sapply(
  1:length(D),
  function (i) inv_logit(aT[,i] + post$bPS[,P[i],S[i]] + post$bD[,P[i]] * D[i]))
diff_ap <- sim_preda - sim_predp # no need to use inverse logit...
dens(diff_ap, lwd = 2, xlab = "Counterfactual predators A-P contrast")
```

Now this actually doesn't look so different from my approach. My only mistake was to use the inverse logit function repeatedly, but the logic of assigning all the tanks to each condition seems to produce the same results as the sampling approach.