---
title: "Week 5 Homework"
author: "Yurun (Ellen) Ying"
date: '2022-06-18'
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(rethinking); library(dagitty)
```

## Course homework

### Problem 1

The total causal effect of gender on grant awards in the `NWOGrants` data.

```{r}
data("NWOGrants")
d <- NWOGrants

# a DAG
dag1 <- dagitty("dag{ G -> D -> A; G -> A}")
coordinates(dag1) <- list(x = c(G = 0, D = 0.5, A = 1),
                          y = c(G = 1, D = 0.3, A = 1))
drawdag(dag1)
```

We will use this model to estimate the total effect of gender on grant awards:

$A_i \sim \mathrm{Binomial}(N_i, p_i)$

$\mathrm{logit}(p_i) = a_{GID[i]}$

$a_{GID[i]} \sim \mathrm{TBD}$

```{r}
# choose priors
# n <- 1e5
# a <- rnorm(n, 0, 1)
# dens(inv_logit(a))

# clean data
dat <- data.frame(
  D = rep(1:9, each = 2),
  G = rep(1:2, 9), # 1 is male, 2 is female
  N = d$applications,
  A = d$awards
)

# fit a model
m1 <- quap(
  alist(
    A ~ dbinom(N, p),
    logit(p) <- a[G],
    a[G] ~ dnorm(0, 1)
  ), data = dat
)

precis(m1, depth = 2)

# posterior prediction
post <- extract.samples(m1)
diff <- inv_logit(post$a[,1]) - inv_logit(post$a[,2])
dens(diff, lwd = 3, xlab = "M-F posterior contrast")
abline(v=0, lty = 2, col = "tomato")
```

Across disciplines, women have a slight disadvantage than men in winning the grant awards.


### Problem 2

Estimate the direct causal effect of gender on grant awards.

```{r}
m2 <- ulam(
  alist(
    A ~ dbinom(N, p),
    logit(p) <- a[G,D],
    matrix[G,D]:a ~ dnorm(0, 1)
  ), data = dat, chains = 4, cores = 4
)

precis(m2, depth = 3)
```

Plot for each discipline

```{r}
post2 <- extract.samples(m2)
PrA <- post2$a
diff_each_D <- sapply(1:9, function (i) inv_logit(PrA[,1,i]) - inv_logit(PrA[,2,i]))

plot(NULL, xlim = c(-0.5, 0.4), ylim = c(0, 18),
     xlab = "M-F posterior contrast", ylab = "Density")
disc <- as.character(d$discipline)
for (i in 1:9) {
  dens(diff_each_D[,i], col = 1 + i, lwd = 4, add = TRUE)
  xloc <- ifelse(mean(diff_each_D[,i]) < 0, -0.35, 0.35)
  xpos <- ifelse(mean(diff_each_D[,i]) < 0, 4, 2)
  text(xloc + 0.5*mean(diff_each_D[,i]), 18 - 1.3*i, disc[2*i], col = 1+i, pos = xpos, font = 2)
}
abline(v = 0, lty = 2, lwd = 2)
```

There are two disciplines where men and women have a roughly same probability of winning awards. In three departments, men are slightly more likely than women to win. In four departments, women are slightly more likely to win. 

Let's now calculate the marginal effect of manipulating (the perception of) applicants' gender.

```{r}
# number of applications in the dataset
total_app <- sum(dat$N)

# number of applications in each discplines
disc_num <- sapply(1:9, function(i) sum(dat$N[dat$D == i]))

# extract samples as if all applicants were men
PrM <- link(m2, 
            data = list(N = rep(1, total_app),
                        D = rep(1:9, times = disc_num),
                        G = rep(1, total_app)))

# extract samples as if all applicants were men
PrF <- link(m2, 
            data = list(N = rep(1, total_app),
                        D = rep(1:9, times = disc_num),
                        G = rep(2, total_app)))

# contrast
diff_marg <- PrM - PrF
mean(diff_marg); PI(diff_marg)
dens(diff_marg, lwd = 2, 
     xlab = "Marginal effect of gender across disciplines", ylab = "Density")
abline(v = 0, lty = 2)
```

The average effect of gender across discipline is slim, but there seems to be heterogeneity of gender difference in different departments.


## Problem 3

There may be more total applicants from those "discriminatory" disciplines which have more weights in contributing to the total effect. Therefore, the total effect turns out to show a slightly higher probability for men to win the grant awards.

### Some nuances

- The disciplines that influence the overall effect the most are those popular disciplines with many applicants
- Those discipline happen to have low award rates
- A large proportion of women applicants is from these disciplines
- In the most popular discipline, more men than women win the award

So both the question of who and where they apply influences the overall effect. This is the G -> D causal path in the DAG.


## Problem 4

The over-representation of left-handers among UFC fighters. Estimate the average advantage, that a left-handed fighter has against right-handed fighter.

```{r}
data("UFClefties")
d <- UFClefties
```

The probability of fight1 winning depends upon the left-handedness of both F1 and F2. There are four possibilities;

1. F1 - R, F2 - R (1)
2. F1 - L, F2 - R (2)
3. F1 - R, F2 - L (3)
4. F1 - L, F2 - L (4)

We will use this model:

$W_i \sim \mathrm{Bernoulli}(p_i)$

$\mathrm{logit}(p_i) = \alpha_{condition[i]}$

$\alpha_{fighter[i]} \sim \mathrm{Normal}(0,1)$

```{r}
# clean dat
d$cid <- d$fighter1.lefty + 2*d$fighter2.lefty + 1
dat <- list(
  W = d$fighter1.win,
  cid = as.integer(d$cid)
)

# fit a model
m3 <- ulam(
  alist(
    W ~ dbinom(1, p),
    logit(p) <- a[cid],
    a[cid] ~ dnorm(0, 1)
  ), data = dat, chains = 4, cores = 4
)

precis(m3, 2)
```

We will do some posterior contrast between condition 1 and 2 as well as between condition 3 and 4.

```{r}
post <- extract.samples(m3)
# advantage of L when opponent is R
diff_12 <- inv_logit(post$a[,2]) - inv_logit(post$a[,1])
# advantage of L when opponent is L
diff_34 <- inv_logit(post$a[,4]) - inv_logit(post$a[,3])
plot(precis(data.frame(diff_12, diff_34)))
```

Left-handers do not seem to have an advantage over right-handers. They even seem to have a slight disadvantage than right-handers.

Estimate the average advantage

```{r}
f2 <- d$fighter2.lefty
# posterior distribution of p as if all fighters are right-handed
f1r_cid <- rep(0, nrow(d)) + 2*f2 + 1
f1r <- link(m3, data = data.frame(cid = f1r_cid))
# posterior distribution of p as if all fighters are left-handed
f1l_cid <- rep(1, nrow(d)) + 2*f2 + 1
f1l <- link(m3, data = data.frame(cid = f1l_cid))

rl_contrast <- f1l - f1r
dens(rl_contrast, xlab = "L-R posterior contrast of winning probability", ylab = "Density")
abline(v = 0, lty = 2)
```

We can see the counterfactual prediction also shows there is no difference between left-handers and right-handers in the probability to win. Left-handers are even slightly disadvantaged.

Left-handers are more represented among fighters probably because they are more visible in such occupations where one handedness is critical and attracts more attention?


### A more legit answer

**The first task - modelling the lefty advantage:** The critical point here is to have a model that applies no matter which fighter is the fighter one. This means that when we hypothetically switch the players, the winning probability needs to be reversed in the model to reflect with switch. Therefore, we formulate the model as following:

$\mathrm{logit}(p_i) = \alpha_{F1[i]} - \alpha_{F2[i]}$

Expand this to include left-handedness:

$\mathrm{logit}(p_i) = (\alpha_{F1[i]} + \beta L_{F1[i]}) - (\alpha_{F2[i]} + \beta L_{F2[i]})$

where $\beta$ is the advantage a lefty will have over a righty. When both fighters are lefties, the second terms in the tw parentheses will cancel out.

Since there is no observation for ability, we fit a model that only includes left-handedness.

```{r}
dat <- list(
  W = d$fighter1.win,
  F1 = d$fighter1.lefty,
  F2 = d$fighter2.lefty
)

m4n <- ulam(
  alist(
    W ~ bernoulli(p),
    logit(p) <- b*(F1 - F2),
    b ~ dnorm(0, 1)
  ), data = dat, chains = 4, cores = 4
)

post <- extract.samples(m4n)
dens(inv_logit(post$b), xlab = "Lefty winning probability when fighting a righty", ylab = "Density")
abline(v = 0.5, lty = 2)
```

This is the distribution of a lefty's winning probability when the opponent is a righty (lefty advantage). The mean is 0.5, so there doesn't seem to be evidence for lefty advantage.

**The second task - explaining the results:** The critical thing here is that we are observing among UFC qualified fighters. This means there is a potential collider bias in our observation.

```{r}
dag2 <- dagitty("dag{ Q <- A -> W; Q <- L -> W}")
coordinates(dag2) <- list(x = c(Q = 0, A = 0.5, W = 1, L = 0.5),
                          y = c(Q = 1, A = 0, W = 1, L = 1))
drawdag(dag2)
```

Only observing qualified fighters is equal to stratifying by Q, which can create a nonexistent association between A and L. It is likely that among the qualified fighters, lefties are on average less able than righties. In the observed matches, the lefty advantage is canceled out by the higher average ability of righties.

We will use some simulations to illustrate this idea.

```{r}
# simulate 5000 fighters
set.seed(89345)
N <- 5000
L <- rbern(N, 0.1)
A <- rnorm(N)
Q <- rep(1, N)

# qualify if A large enough or lefty
Q <- ifelse(A > 2 | (A > 1.25 & L == 1), 1, 0)

# summarize
table(Q, L)
```

Check the proportion of lefty among the qualified fighters

```{r}
sum(Q == 1 & L == 1)/sum(Q == 1)
```

Nearly 39% of qualified fighters are lefty, which is way higher than the baseline 10%.

The mean ability of lefties and righties among the qualified fighter

```{r}
# righty
mean(A[Q == 1 & L == 0])
# lefty
mean(A[Q == 1 & L == 1])
```

Lefties' mean ability is lower than righties'.

Now simulate the fight result

```{r}
b <- 0.5 # lefty advantage
k <- 2 # importance of ability difference
# get the ability and left-handedness of qualified fighters
l <- L[Q == 1]
a <- A[Q == 1]
M <- sum(Q==1)/2 # number of matches
f1win <- rep(0, M)
# record the left-handedness of the two fighters
L1 <- rep(0, M); L2 <- rep(0, M)

# simulate the fight
for (i in 1:M) {
  # a1 is from the first half of all fighters, a2 is from the second half
  a1 <- a[i] + b*l[i]
  a2 <- a[i + M] + b*l[i + M]
  p_f1win <- inv_logit(k*(a1 - a2))
  f1win[i] <- rbern(1, prob = p_f1win)
  L1[i] <- l[i]
  L2[i] <- l[i+M]
}
```

Fit a model to the simulated dataset

```{r}
datx <- list(W = f1win, F1 = L1, F2 = L2)

m4x <- ulam(
  alist(
    W ~ bernoulli(p),
    logit(p) <- b*(F1 - F2),
    b ~ dnorm(0, 1)
  ), data = datx, chains = 4, cores = 4
)

post <- extract.samples(m4x)
dens(inv_logit(post$b), xlab = "Lefty winning probability when fighting a righty", ylab = "Density")
abline(v = 0.5, lty = 2)
```

Now a lefty in fact has a lower probability winning a righty. This probability will increase when lefty advantage is stronger, or/and when the qualification for lefty is higher.


## Problems from the book

### 11H1

Compare the predictive accuracy of different models of prosocial chimps.

```{r, include=FALSE}
data("chimpanzees")
d <- chimpanzees
# coding the four treatment conditions
d$treatment <- 1 + d$prosoc_left + 2*d$condition

dat_list <- list(
  pulled_left = d$pulled_left,
  actor = d$actor,
  treatment = as.integer(d$treatment)
)
```

```{r}
# model with a unique intercept for each actor
m11.4 <- ulam(
  alist(
    pulled_left ~ dbinom(1, p),
    logit(p) <- a[actor] + b[treatment],
    a[actor] ~ dnorm(0, 10),
    b[treatment] ~ dnorm(0, 0.5)
  ),
  data = dat_list, chains = 4, log_lik = TRUE
)

# simpler model
m11.1 <- ulam(
    alist(
      pulled_left ~ dbinom(1, p),
      logit(p) <- a,
      a ~ dnorm(0, 1.5)
    ),
    data = dat_list, chains = 4, log_lik = TRUE
  )

m11.2 <- ulam(
    alist(
      pulled_left ~ dbinom(1, p),
      logit(p) <- a + b[treatment],
      a ~ dnorm(0, 1.5),
      b[treatment] ~ dnorm(0, 0.5)
    ),
    data = dat_list, chains = 4, log_lik = TRUE
  )

compare(m11.1, m11.2, m11.4, func = PSIS)
```

The model including a unique intercept for each actor does better than both the models with only an intercept or the model with an intercept and treatment as a predictor. This means that both each actor's baseline probability and the treatment are important in predicting the probability of pulling the lever.


### 11H2

Salmon pirating attempts by Bal Eagles in Washington State data.

```{r}
library(MASS);data(eagles)
d <- eagles
d$P <- as.integer(ifelse(d$P == "L", 1, 0))
d$V <- as.integer(ifelse(d$V == "L", 1, 0))
d$A <- as.integer(ifelse(d$A == "A", 1, 0))
```

(a) Fit the following model:

$y_i \sim \mathrm{Binomial}(n_i, p_i)$

$\mathrm{logit}(p_i) = \alpha + \beta_P P_i + \beta_V V_i + \beta_A A_i$

$\alpha \sim \mathrm{Normal}(0, 1.5)$

$\beta_P, \beta_V, \beta_A \sim \mathrm{Normal}(0, 0.5)$

where y is the number of successful attempts, n is the total number of attempts, P is a dummy variable indicating whether or not the pirate had large body size, V is a dummy variable indicating whether or not the victim had large body size, and finally A is a dummy variable indicating whether or not the pirate was an adult. 

```{r}
# a quap model
m11h2_quap <- quap(
  alist(
    y ~ dbinom(n, p),
    logit(p) <- a + bP*P + bV*V + bA*A,
    a ~ dnorm(0, 1.5),
    c(bP,bV,bA) ~ dnorm(0, 0.5)
  ), data = d
)

precis(m11h2_quap)
```

```{r}
# a ulam model
m11h2_ulam <- ulam(
  alist(
    y ~ dbinom(n, p),
    logit(p) <- a + bP*P + bV*V + bA*A,
    a ~ dnorm(0, 1.5),
    c(bP,bV,bA) ~ dnorm(0, 0.5)
  ), data = d, chains = 4, cores = 4, log_lik = TRUE
)

precis(m11h2_ulam)
```

Both models seem to give good estimations.

(b) Plot the posterior predictions. Compute and display both (1) the predicted **probability** of success and its 89% interval for each row (i) in the data, as well as (2) the predicted success **count** and its 89% interval. What different information does each type of posterior prediction provide?

Posterior prediction of success probability

```{r}
set.seed(2798)
post_p <- link(m11h2_ulam, 
               data = data.frame(n = d$n,
                                 P = d$P,
                                 A = d$A,
                                 V = d$V))

par(mfrow = c(1, 2))
# pirate adult
post_p_A1 <- post_p[, which(d$A == 1)] 
mean_p_A1 <- apply(post_p_A1, 2, mean)
PI_p_A1 <- apply(post_p_A1, 2, PI)
plot(NULL,
     xlab = "Conditions", ylab = "Predicted expectation of probability",
     xlim = c(1,4), ylim = c(0,1), xaxt = 'n')
for (i in 1:4) lines(c(i,i), PI_p_A1[,i], lwd = 2)
points(1:4, mean_p_A1, pch = 16, cex = 1.5, col = "white")
points(1:4, mean_p_A1, pch = 16, cex = 1, col = "black")
axis(side = 1, at = 1:4, 
     labels = c("PL/VL", "PL/VS", "PS/VL", "PS/VS"))
mtext("Pirate is an adult")

# pirate immature
post_p_A0 <- post_p[, which(d$A == 0)] 
mean_p_A0 <- apply(post_p_A0, 2, mean)
PI_p_A0 <- apply(post_p_A0, 2, PI)
plot(NULL,
     xlab = "Conditions", ylab = "Predicted expectation of probability",
     xlim = c(1,4), ylim = c(0,1), xaxt = 'n')
for (i in 1:4) lines(c(i,i), PI_p_A0[,i], lwd = 2)
points(1:4, mean_p_A0, pch = 16, cex = 1.5, col = "white")
points(1:4, mean_p_A0, pch = 16, cex = 1, col = "black")
axis(side = 1, at = 1:4, 
     labels = c("PL/VL", "PL/VS", "PS/VL", "PS/VS"))
mtext("Pirate is immature")
```


Posterior prediction of success count

```{r}
set.seed(2798)
post_c <- sim(m11h2_ulam, 
               data = data.frame(n = rep(30, nrow(d)),
                                 P = d$P,
                                 A = d$A,
                                 V = d$V))

par(mfrow = c(1, 2))
# pirate adult
post_c_A1 <- post_c[, which(d$A == 1)] 
mean_c_A1 <- apply(post_c_A1, 2, mean)
PI_c_A1 <- apply(post_c_A1, 2, PI)
plot(NULL,
     xlab = "Conditions", ylab = "Predicted expectation of count (max=30)",
     xlim = c(1,4), ylim = c(0,30), xaxt = 'n')
for (i in 1:4) lines(c(i,i), PI_c_A1[,i], lwd = 2)
points(1:4, mean_c_A1, pch = 16, cex = 1.5, col = "white")
points(1:4, mean_c_A1, pch = 16, cex = 1, col = "black")
axis(side = 1, at = 1:4, 
     labels = c("PL/VL", "PL/VS", "PS/VL", "PS/VS"))
mtext("Pirate is an adult")

# pirate immature
post_c_A0 <- post_c[, which(d$A == 0)] 
mean_c_A0 <- apply(post_c_A0, 2, mean)
PI_c_A0 <- apply(post_c_A0, 2, PI)
plot(NULL,
     xlab = "Conditions", ylab = "Predicted expectation of count (max=30)",
     xlim = c(1,4), ylim = c(0,30), xaxt = 'n')
for (i in 1:4) lines(c(i,i), PI_c_A0[,i], lwd = 2)
points(1:4, mean_c_A0, pch = 16, cex = 1.5, col = "white")
points(1:4, mean_c_A0, pch = 16, cex = 1, col = "black")
axis(side = 1, at = 1:4, 
     labels = c("PL/VL", "PL/VS", "PS/VL", "PS/VS"))
mtext("Pirate is immature")
```


The prediction of probability contains uncertainty from the intercepts and the coefficients in the model. The prediction of count also contains the uncertainty in generating the count data from the probability via a biomial distribution. The count predication contains more uncertainty.


(c) Now try to improve the model. Consider an interaction between the pirate’s size and age (immature or adult). Compare this model to the previous one, using WAIC. Interpret.

```{r}
m11h2_ulam_n <- ulam(
  alist(
    y ~ binomial(n, p),
   logit(p) <- a + bP*P + bV*V + bA*A + bPA*P*A,
    a ~ dnorm(0, 1.5),
    c(bP,bV,bA,bPA) ~ dnorm(0, 0.5)
  ), data = d, chains = 4, cores = 4, log_lik = TRUE
)

compare(m11h2_ulam, m11h2_ulam_n, func = WAIC)
```


The new model containing an interaction term works worse than the original model.


## 11H6

Modeling the association between social learning and brain size.

```{r}
data("Primates301")
d <- Primates301
# a clean data
dat <- data.frame(
  L = d$social_learning,
  log_b = log(d$brain)
)
# complete cases
dat <- dat[complete.cases(dat$L, dat$log_b),]
dat$B <- dat$log_b/max(dat$log_b) # standardize brain size
```

(a) Model the number of observations of `social_learning` for each species as a function of the log `brain size`. Use a Poisson distribution for the `social_learning` outcome variable. Interpret the resulting posterior. 

The model we will be using

$L_i \sim \mathrm{Poisson}(\lambda_i)$

$\mathrm{log}(\lambda_i) = \alpha + \beta B_i$

```{r}
# choose priors
N <- 100
a <- rnorm(N, 0, 1)
b <- rnorm(N, 0, 1)
plot(NULL, xlim = c(0, 1), ylim = c(0, 10), 
     xlab = "Log brain size (std)", ylab = "Social learning")
for (i in 1:N) curve(exp(a[i]+b[i]*x), from = 0, to = 1, 
                     col = col.alpha("black", 0.3), add = TRUE)

```

```{r}
# fit a model
m11h6a <- ulam(
  alist(
    L ~ dpois(lambda),
    log(lambda) <- a + b*B,
    a ~ dnorm(0, 1),
    b ~ dnorm(0, 1)
  ), data = dat, chains = 4, cores = 4
)

precis(m11h6a)

# posterior prediction
B_seq <- seq(0, 1, length.out = 50)
post_l <- link(m11h6a, data = data.frame(B = B_seq))
l_mu <- apply(post_l, 2, mean)
l_PI <- apply(post_l, 2, PI)
plot(NULL, xlim = c(0, 1), ylim = c(0, 220),
     xlab = "Log brain size (std)", ylab = "Social learning")
points(dat$B, dat$L, col = "turquoise")
lines(B_seq, l_mu)
shade(l_PI, B_seq)

# on a natural scale
B_natural <- exp(B_seq*max(dat$log_b))
plot(exp(dat$log_b), dat$L, col = "turquoise", 
     xlim =range(B_natural), ylim = c(0, 220),
     xlab = "Brain size", ylab = "Social learning")
lines(B_natural, l_mu)
shade(l_PI, B_natural)
```
This model indicates that social learning grows exponentially with brain size.

(b) Some species are studied much more than others. So the number of reported instances of `social_learning` could be a product of research effort. Use the `research_effort` variable, specifically its logarithm, as an additional predictor variable. Interpret the coefficient for log `research_effort`. How does this model differ from the previous one?

```{r}
# add variables
dat2 <- data.frame(
  L = d$social_learning,
  log_b = log(d$brain),
  log_r = log(d$research_effort)
)
dat2 <- dat2[complete.cases(dat2$L, dat2$log_b, dat2$log_r),]
dat2$B <- dat2$log_b/max(dat2$log_b) # standardize brain size
dat2$R <- dat2$log_r/max(dat2$log_r)
```

```{r}
# fit a model
m11h6b <- ulam(
  alist(
    L ~ dpois(lambda),
    log(lambda) <- a + bB*B + bR*R,
    a ~ dnorm(0, 1),
    c(bB,bR) ~ dnorm(0, 1)
  ), data = dat2, chains = 4, cores = 4
)

precis(m11h6b)

exp(coef(m11h6b))
```

One unit increase in log brain size would result in an increase of 11 social learning counts. This is a more conservative estimate of the effect of brain size on social learning than the model only containing brain size as a predictor.

(c) Draw a DAG to represent how you think the variables `social_learning`, `brain`, and `research_effort` interact. Justify the DAG with the measured associations in the two models above (and any other models you used).

```{r}
dag3 <- dagitty("dag{ 
                U[unobserved]
                B <- U -> R; R -> L; B -> L}")
coordinates(dag3) <- list(x = c(U = 0, B = 0, R = 1, L = 1),
                          y = c(U = 0, B = 1, R = 0, L = 1))
drawdag(dag3)
```

Stratifying by research effort reduces the causal effect of log brain size indicates that there might be a fork structure in the causal model. It is impossible for research effort to cause bigger brain, so there might be other variables causing both brain size and research effort. This might be the closeness to human being. The closeness can make species have similar big brain to homo sapiens, and it can also cause more curiosity to study them, resulting in more research effort in studying these species.