---
title: "Week 5 Homework"
author: "Yurun (Ellen) Ying"
date: '2022-06-18'
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(rethinking); library(dagitty)
```

## Course homework

### Problem 1

The total causal effect of gender on grant awards in the `NWOGrants` data.

```{r}
data("NWOGrants")
d <- NWOGrants

# a DAG
dag1 <- dagitty("dag{ G -> D -> A; G -> A}")
coordinates(dag1) <- list(x = c(G = 0, D = 0.5, A = 1),
                          y = c(G = 1, D = 0.3, A = 1))
drawdag(dag1)
```

We will use this model to estimate the total effect of gender on grant awards:

$A_i \sim \mathrm{Binomial}(N_i, p_i)$

$\mathrm{logit}(p_i) = a_{GID[i]}$

$a_{GID[i]} \sim \mathrm{TBD}$

```{r}
# choose priors
# n <- 1e5
# a <- rnorm(n, 0, 1)
# dens(inv_logit(a))

# clean data
dat <- data.frame(
  D = rep(1:9, each = 2),
  G = rep(1:2, 9), # 1 is male, 2 is female
  N = d$applications,
  A = d$awards
)

# fit a model
m1 <- quap(
  alist(
    A ~ dbinom(N, p),
    logit(p) <- a[G],
    a[G] ~ dnorm(0, 1)
  ), data = dat
)

precis(m1, depth = 2)

# posterior prediction
post <- extract.samples(m1)
diff <- inv_logit(post$a[,1]) - inv_logit(post$a[,2])
dens(diff, lwd = 3, xlab = "M-F posterior contrast")
abline(v=0, lty = 2, col = "tomato")
```

Across disciplines, women have a slight disadvantage than men in winning the grant awards.


### Problem 2

Estimate the direct causal effect of gender on grant awards.

```{r}
m2 <- ulam(
  alist(
    A ~ dbinom(N, p),
    logit(p) <- a[G,D],
    matrix[G,D]:a ~ dnorm(0, 1)
  ), data = dat, chains = 4, cores = 4
)

precis(m2, depth = 3)
```

Plot for each discipline

```{r}
post2 <- extract.samples(m2)
PrA <- post2$a
diff_each_D <- sapply(1:9, function (i) inv_logit(PrA[,1,i]) - inv_logit(PrA[,2,i]))

plot(NULL, xlim = c(-0.5, 0.4), ylim = c(0, 18),
     xlab = "M-F posterior contrast", ylab = "Density")
disc <- as.character(d$discipline)
for (i in 1:9) {
  dens(diff_each_D[,i], col = 1 + i, lwd = 4, add = TRUE)
  xloc <- ifelse(mean(diff_each_D[,i]) < 0, -0.35, 0.35)
  xpos <- ifelse(mean(diff_each_D[,i]) < 0, 4, 2)
  text(xloc + 0.5*mean(diff_each_D[,i]), 18 - 1.3*i, disc[2*i], col = 1+i, pos = xpos, font = 2)
}
abline(v = 0, lty = 2, lwd = 2)
```

There are two disciplines where men and women have a roughly same probability of winning awards. In three departments, men are slightly more likely than women to win. In four departments, women are slightly more likely to win. 

Let's now calculate the marginal effect of manipulating (the perception of) applicants' gender.

```{r}
# number of applications in the dataset
total_app <- sum(dat$N)

# number of applications in each discplines
disc_num <- sapply(1:9, function(i) sum(dat$N[dat$D == i]))

# extract samples as if all applicants were men
PrM <- link(m2, 
            data = list(N = rep(1, total_app),
                        D = rep(1:9, times = disc_num),
                        G = rep(1, total_app)))

# extract samples as if all applicants were men
PrF <- link(m2, 
            data = list(N = rep(1, total_app),
                        D = rep(1:9, times = disc_num),
                        G = rep(2, total_app)))

# contrast
diff_marg <- PrM - PrF
mean(diff_marg); PI(diff_marg)
dens(diff_marg, lwd = 2, 
     xlab = "Marginal effect of gender across disciplines", ylab = "Density")
abline(v = 0, lty = 2)
```

The average effect of gender across discipline is slim, but there seems to be heterogeneity of gender difference in different departments.


## Problem 3

There may be more total applicants from those "discriminatory" disciplines which have more weights in contributing to the total effect. Therefore, the total effect turns out to show a slightly higher probability for men to win the grant awards.

### Some nuances

- The disciplines that influence the overall effect the most are those popular disciplines with many applicants
- Those discipline happen to have low award rates
- A large proportion of women applicants is from these disciplines
- In the most popular discipline, more men than women win the award

So both the question of who and where they apply influences the overall effect. This is the G -> D causal path in the DAG.


## Problem 4

The over-representation of left-handers among UFC fighters. Estimate the average advantage, that a left-handed fighter has against right-handed fighter.

```{r}
data("UFClefties")
d <- UFClefties
```

The probability of fight1 winning depends upon the left-handedness of both F1 and F2. There are four possibilities;

1. F1 - R, F2 - R (1)
2. F1 - L, F2 - R (2)
3. F1 - R, F2 - L (3)
4. F1 - L, F2 - L (4)

We will use this model:

$W_i \sim \mathrm{Bernoulli}(p_i)$

$\mathrm{logit}(p_i) = \alpha_{condition[i]}$

$\alpha_{fighter[i]} \sim \mathrm{Normal}(0,1)$

```{r}
# clean dat
d$cid <- d$fighter1.lefty + 2*d$fighter2.lefty + 1
dat <- list(
  W = d$fighter1.win,
  cid = as.integer(d$cid)
)

# fit a model
m3 <- ulam(
  alist(
    W ~ dbinom(1, p),
    logit(p) <- a[cid],
    a[cid] ~ dnorm(0, 1)
  ), data = dat, chains = 4, cores = 4
)

precis(m3, 2)
```

We will do some posterior contrast between condition 1 and 2 as well as between condition 3 and 4.

```{r}
post <- extract.samples(m3)
# advantage of L when opponent is R
diff_12 <- inv_logit(post$a[,2]) - inv_logit(post$a[,1])
# advantage of L when opponent is L
diff_34 <- inv_logit(post$a[,4]) - inv_logit(post$a[,3])
plot(precis(data.frame(diff_12, diff_34)))
```

Left-handers do not seem to have an advantage over right-handers. They even seem to have a slight disadvantage than right-handers.

Estimate the average advantage

```{r}
f2 <- d$fighter2.lefty
# posterior distribution of p as if all fighters are right-handed
f1r_cid <- rep(0, nrow(d)) + 2*f2 + 1
f1r <- link(m3, data = data.frame(cid = f1r_cid))
# posterior distribution of p as if all fighters are left-handed
f1l_cid <- rep(1, nrow(d)) + 2*f2 + 1
f1l <- link(m3, data = data.frame(cid = f1l_cid))

rl_contrast <- f1l - f1r
dens(rl_contrast, xlab = "L-R posterior contrast of winning probability", ylab = "Density")
abline(v = 0, lty = 2)
```

We can see the counterfactual prediction also shows there is no difference between left-handers and right-handers in the probability to win. Left-handers are even slightly disadvantaged.

Left-handers are more represented among fighters probably because they are more visible in such occupations where one handedness is critical and attracts more attention?


### A more legit answer

**The first task - modelling the lefty advantage:** The critical point here is to have a model that applies no matter which fighter is the fighter one. This means that when we hypothetically switch the players, the winning probability needs to be reversed in the model to reflect with switch. Therefore, we formulate the model as following:

$\mathrm{logit}(p_i) = \alpha_{F1[i]} - \alpha_{F2[i]}$

Expand this to include left-handedness:

$\mathrm{logit}(p_i) = (\alpha_{F1[i]} + \beta L_{F1[i]}) - (\alpha_{F2[i]} + \beta L_{F2[i]})$

where $\beta$ is the advantage a lefty will have over a righty. When both fighters are lefties, the second terms in the tw parentheses will cancel out.

Since there is no observation for ability, we fit a model that only includes left-handedness.

```{r}
dat <- list(
  W = d$fighter1.win,
  F1 = d$fighter1.lefty,
  F2 = d$fighter2.lefty
)

m4n <- ulam(
  alist(
    W ~ bernoulli(p),
    logit(p) <- b*(F1 - F2),
    b ~ dnorm(0, 1)
  ), data = dat, chains = 4, cores = 4
)

post <- extract.samples(m4n)
dens(inv_logit(post$b), xlab = "Lefty winning probability when fighting a righty", ylab = "Density")
abline(v = 0.5, lty = 2)
```

This is the distribution of a lefty's winning probability when the opponent is a righty (lefty advantage). The mean is 0.5, so there doesn't seem to be evidence for lefty advantage.

**The second task - explaining the results:** The critical thing here is that we are observing among UFC qualified fighters. This means there is a potential collider bias in our observation.

```{r}
dag2 <- dagitty("dag{ Q <- A -> W; Q <- L -> W}")
coordinates(dag2) <- list(x = c(Q = 0, A = 0.5, W = 1, L = 0.5),
                          y = c(Q = 1, A = 0, W = 1, L = 1))
drawdag(dag2)
```

Only observing qualified fighters is equal to stratifying by Q, which can create a nonexistent association between A and L. It is likely that among the qualified fighters, lefties are on average less able than righties. In the observed matches, the lefty advantage is canceled out by the higher average ability of righties.

We will use some simulations to illustrate this idea.

```{r}
# simulate 5000 fighters
set.seed(89345)
N <- 5000
L <- rbern(N, 0.1)
A <- rnorm(N)
Q <- rep(1, N)

# qualify if A large enough or lefty
Q <- ifelse(A > 2 | (A > 1.25 & L == 1), 1, 0)

# summarize
table(Q, L)
```

Check the proportion of lefty among the qualified fighters

```{r}
sum(Q == 1 & L == 1)/sum(Q == 1)
```

Nearly 39% of qualified fighters are lefty, which is way higher than the baseline 10%.

The mean ability of lefties and righties among the qualified fighter

```{r}
# righty
mean(A[Q == 1 & L == 0])
# lefty
mean(A[Q == 1 & L == 1])
```

Lefties' mean ability is lower than righties'.

Now simulate the fight result

```{r}
b <- 0.5 # lefty advantage
k <- 2 # importance of ability difference
# get the ability and left-handedness of qualified fighters
l <- L[Q == 1]
a <- A[Q == 1]
M <- sum(Q==1)/2 # number of matches
f1win <- rep(0, M)
# record the left-handedness of the two fighters
L1 <- rep(0, M); L2 <- rep(0, M)

# simulate the fight
for (i in 1:M) {
  # a1 is from the first half of all fighters, a2 is from the second half
  a1 <- a[i] + b*l[i]
  a2 <- a[i + M] + b*l[i + M]
  p_f1win <- inv_logit(k*(a1 - a2))
  f1win[i] <- rbern(1, prob = p_f1win)
  L1[i] <- l[i]
  L2[i] <- l[i+M]
}
```

Fit a model to the simulated dataset

```{r}
datx <- list(W = f1win, F1 = L1, F2 = L2)

m4x <- ulam(
  alist(
    W ~ bernoulli(p),
    logit(p) <- b*(F1 - F2),
    b ~ dnorm(0, 1)
  ), data = datx, chains = 4, cores = 4
)

post <- extract.samples(m4x)
dens(inv_logit(post$b), xlab = "Lefty winning probability when fighting a righty", ylab = "Density")
abline(v = 0.5, lty = 2)
```

Now a lefty in fact has a lower probability winning a righty. This probability will increase when lefty advantage is stronger, or/and when the qualification for lefty is higher.