---
title: 'Lecture 14.5: Instrumental Variables'
author: "Yurun (Ellen) Ying"
date: '2022-07-04'
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(rethinking); library(dagitty)
```

## Instrumental variables

Correlated varying effects can also be implemented in constructing instrumental variables for causal inferences. 

Instrumental variables have to satisfy the following three criteria:

- Independent of confounder U
- Not independent of exposure X
- Cannot influence outcome Y except through X

An example is th effect of education on wages. There are many confounders, and we choose the season of birth as the instrumental variable.

```{r}
dag1 <- dagitty("dag{U[unobserved]; Q -> E -> W; E <- U-> W}")
coordinates(dag1) <- list(x = c(Q = 0.3, U = 0.5, E = 0.4, W = 0.6),
                          y = c(Q = 0.25, U = 0.25, E = 0.75, W = 0.75))
drawdag(dag1)
```

A generative model of the data

```{r}
set.seed(73)
N <- 500
U <- rnorm(N)
Q <- sample(1:4, size = N, replace = TRUE)
E <- rnorm(N, U + Q)
W <- rnorm(N, U) # education doesn't affect wages
dat_sim <- list(W = standardize(W),
                E = standardize(E),
                Q = standardize(Q))
```

The way to model this is just to write a statistical version of generative model. To capture the effect of the unobserved confounders, we will model E and W as from the same multivariate normal distribution:

$$\begin{aligned}
\begin{pmatrix}
W_i\\
E_i
\end{pmatrix} &\sim \mathrm{MVNormal}(
\begin{pmatrix}
\mu_{W,i}\\
\mu_{E,i}
\end{pmatrix}, 
\mathbf{S}
) \\
\mu_{W,i} &= \alpha_W + \beta_{EW} E_i \\
\mu_{E,i} &= \alpha_E + \beta_{QE}Q_i
\end{aligned}$$

```{r, results="hide"}
m <- ulam(
  alist(
    c(W,E) ~ multi_normal(c(muW,muE), Rho, Sigma),
    muW <- aW + bEW*E,
    muE <- aE + bQE*Q,
    c(aW,aE) ~ normal(0, 0.2),
    c(bEW,bQE) ~ normal(0, 0.5),
    Rho ~ lkj_corr(2),
    Sigma ~ exponential(1)
  ), data = dat_sim, chains = 4, cores = 4
)
```

```{r}
round(precis(m, depth = 3), 2)
```


The influence of E on W is around zero, which reflects the true causal influence. The correlation between E and W is reliably positive, reflecting the common cause U.