---
title: 'Lecture 3: Geocentric Models'
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Examples of normal distribution

Normal by addition: any process that adds value from the same distribution converges to a normal distribution,
```{r}
pos <- replicate(1000, sum(runif(16, -1, 1)))
plot(density(pos))
```

 Normal by multiplication: multiplication of small values converges to normal.
```{r}
pos.small <- replicate(1000, prod(1 + runif(12, 0, 0.01)))
plot(density(pos.small))

# doesn't work for big value
pos.big <- replicate(1000, prod(1 + runif(12, 0, 0.5)))
plot(density(pos.big))

# but still works when we take the log of the product
# since it's the same as add them together
log.big <- replicate(1000, log(prod(1 + runif(12, 0, 0.5))))
plot(density(log.big))
```
 
## Build a Gaussian model for height

Build a linear model of height, the following is our prior:

- h ~ normal(mu, sigma)
- mu ~ normal(178, 20)
- sigma ~ uniform(0, 50)

```{r, message=FALSE}
# read in data
library(rethinking)
data("Howell1")
d <- Howell1
d2 <- d[d$age >= 18, ]
```

Use the model to do prior predicative simulation to see whether the assumptions make sense.
```{r}
n <- 1e4
sample_mu <- rnorm(n, mean = 178, sd = 20)
sample_sigma <- runif(n, min = 0, max = 50)
prior_h <- rnorm(n, mean = sample_mu, sd = sample_sigma)
dens(prior_h)
```

Fit the model using grid approximation.
```{r}
# set the grid and get a list of all combinations of the two parameters
mu.list <- seq(150, 160, length.out = 100)
sigma.list <- seq(7, 9, length.out = 100)
post <- expand.grid(mu = mu.list, sigma = sigma.list)

# a vector of log likelihood of each parameter combinations producing the data
# the prob of each height value are added together because
#   1. assuming each height is independent of each other
#   2. we are taking the log value of the probability
post$LL <- 
  sapply(
    1:nrow(post), 
    function(i) sum(dnorm(d2$height, post$mu[i], post$sigma[i], log = TRUE))
    )

# counting the uncertainty of the parameters
post$prod <- 
  post$LL + 
  dnorm(post$mu, 178, 20, log = TRUE) + 
  dunif(post$sigma, 0, 50, log = TRUE)

# exponentiate to get the probability
# note that this is not the exact probability but relative ones
# this step is necessary because R round small values to zero, and is not helpful when the values are rare
post$prob <- exp(post$prod - max(post$prod))

# plotting
contour_xyz(post$mu, post$sigma, post$prob)
image_xyz(post$mu, post$sigma, post$prob)
```

Use the posterior distribution to do posterior predicative simulation.
```{r}
sample.rows <- sample(1:n, size = n, replace = TRUE, prob = post$prob)
sample.mu <- post$mu[sample.rows]
sample.sigma <- post$sigma[sample.rows]
plot(sample.mu, sample.sigma, cex = 0.5, pch = 16, col = col.alpha(rangi2, 0.2))
```

Check the distribution of parameters independently.
```{r}
dens(sample.mu)
PI(sample.mu)
dens(sample.sigma)
PI(sample.sigma)
```


## Use quadratic approximation to estimate the posterior distribution

The `rethinking` package provides a convenient function `quap` to do quadratic approximation.

```{r}
# we first construct a formula for the function
flist <- alist(
  height ~ dnorm(mu, sigma),
  mu ~ dnorm(178, 20),
  sigma ~ dunif(0, 50)
)

# and then approximate
m4.1 <- quap(flist, data = d2)
precis(m4.1)

# get the variance-covariance matrix
# and parameter variance and correlation matrix
# the correlation between the two parameters is small
# because they are independent form each other
diag(vcov(m4.1)); cov2cor(vcov(m4.1))

# posterior predicative simulation
post.sample <- extract.samples(m4.1, n = n)
head(post.sample)
precis(m4.1)

```

### Inside the brain of extract.samples function

```{r, message=FALSE}
# the mechanism behind this function is the mvrnorm from the MASS package
library(MASS)
post <- mvrnorm(n = 1e4, mu = coef(m4.1), Sigma = vcov(m4.1))
head(post)
```


## A linear prediction model of height by weight

This model does linear prediction of height from weight. We first pick some priors and do prior predicative simulation.
```{r}
# read data
data(Howell1); d <- Howell1; d2 <- d[d$age >= 18, ]
plot(d2$height ~ d2$weight)

# prior predicative distribution
n <- 100
set.seed(2971)
a <- rnorm(n, 178, 20)
b <- rnorm(n, 0, 10)

# plotting
# note that this prior does not make much sense, so we need to change 
plot(NULL, xlim = range(d2$weight), ylim = c(-100, 400),
     xlab = "weight", ylab = "height")
abline(h = 0, lty = 2)
abline(h = 272, lty = 1, lwd = 0.5)
mtext("b ~ dnorm(0, 10)")
xbar <- mean(d2$weight)
for (i in 1:n) curve(a[i] + b[i]*(x - xbar),
        from = min(d2$weight), to = max(d2$weight),
        add = TRUE, col = col.alpha("black", 0.2))

# use log normal distribution
# this is a more reasonable choice of priors
b <- rlnorm(n, 0, 1)
plot(NULL, xlim = range(d2$weight), ylim = c(-100, 400),
     xlab = "weight", ylab = "height")
abline(h = 0, lty = 2)
abline(h = 272, lty = 1, lwd = 0.5)
mtext("log(b) ~ dnorm(0, 10)")
xbar <- mean(d2$weight)
for (i in 1:n) curve(a[i] + b[i]*(x - xbar),
                     from = min(d2$weight), to = max(d2$weight),
                     add = TRUE, col = col.alpha("black", 0.2))
```

Now let's fit the model to the data.
```{r}
m4.3 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <-  a + b * (weight - xbar),
    a ~ dnorm(178, 20),
    b ~ dlnorm(0, 1),
    sigma ~ dunif(0, 50)
  ),
  data = d2
)

# see what the model is doing
# one additional unit of weight adds to the height by 0.9
# the mostly likely value of b lies in 0.84-0.97
precis(m4.3) 

# have little covariance due to centering
round(vcov(m4.3), 3) 

# plot the posterior distribution against data
plot(height ~ weight, data = d2, 
     xlim = range(d2$weight), ylim = range(d2$height),
     col = rangi2)
mtext(paste("N = ", nrow(d2)))
post <- extract.samples(m4.3, n = 20)
for (i in 1:20) curve(
  post$a[i] + post$b[i] * (x - xbar), 
  add = TRUE, col = col.alpha("black", 0.2))

```


### Plotting a regression interval/contour

We first plot the interval of the mean values of height.
```{r}
# define values of weight to make compute predictions for
# these values will be one the horizontal axis
weight.seq <- seq(from = 25, to = 70, by = 1)

# use link to compute mu
# for each value in weight.seq
# for each sample from posterior distribution
mu <- link(m4.3, data = data.frame(weight = weight.seq))
str(mu)

# plot the points
plot(height ~ weight, d2, type = "n")
for (i in 1:1000) points(weight.seq, mu[i, ], 
                        pch = 16, col = col.alpha(rangi2, 0.1))
```

We turn the points into q contour.
```{r}
mu.mean <- apply(mu, 2, mean) # calculate the mean for each column
mu.PI <- apply(mu, 2, PI, prob = 0.89)

# plot the raw data
plot(height ~ weight, data = d2, col = col.alpha(rangi2, 0.5))
# a line for the mean values of mu
lines(weight.seq, mu.mean)
# add the contour
shade(mu.PI, weight.seq)
```

The contour above depicts the uncertainty surrounding the mean values of height. There is also uncertainty in actual height values and we need to plot this as well.
```{r}
# prediction interval for the actual height
# get a collection of simulated height at each value of weigh
sim.height <- sim(m4.3, data = data.frame(weight = weight.seq))
str(sim.height)

# summarize the compatibility interval
height.PI <- apply(sim.height, 2, PI, prob = 0.89)

# add the interval to the previous plot
plot(height ~ weight, data = d2, col = col.alpha(rangi2, 0.5))
lines(weight.seq, mu.mean)
shade(mu.PI, weight.seq)
shade(height.PI, weight.seq)
```

#### Inside the brain of link and sim function

`link`: get samples from the posterior distribution and use them to predict from a sequence of weight value
```{r}
post <- extract.samples(m4.3, n = 1000)
mu.link <- function(weight) post$a + post$b * (weight - xbar)
weight.seq <- seq(from = 25, to = 70, by = 1)
mu <- sapply(weight.seq, mu.link)
```

`sim`: get samples from the posterior distribution and draw height value from normal distributions defined by each parameter combination
```{r}
post <- extract.samples(m4.3, n = 1000)
weight.seq <- seq(from = 25, to = 70, by = 1)
sim <- function(weight) rnorm(
  n = nrow(post),
  mean = post$a + post$b * (weight - xbar),
  sd = post$sigma
)
sim.height <- sapply(weight.seq, sim)
```

