---
title: 'Lecture 13: Multi-Multilevel Model'
author: "Yurun (Ellen) Ying"
date: '2022-06-25'
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(rethinking)
```


## Multiple clusters

We will revisit the chimpanzee data. This time we want to consider both how treatment and block (experiment days) affect the outcome (pulling left lever). We will use a full interaction model:

$P_i \sim \mathrm{Bernoulli}(p_i)$

$\mathrm{logit}(p_i) = \alpha_{A[i]} + \beta_{T[i], B[i]}$

$\alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma_{\alpha})$

$\beta_{j,k} \sim \mathrm{Normal}(0, \sigma_{\beta})$

$\bar{\alpha} \sim \mathrm{Normal}(0, 1.5)$

$\sigma_{\alpha}, \sigma_{\beta} \sim \mathrm{Exponential}(1)$

```{r, include=FALSE}
data("chimpanzees")
d <- chimpanzees
d$treatment <- 1 + d$prosoc_left + 2*d$condition
```

```{r}
# clean data
dat <- list(
  P = d$pulled_left,
  A = d$actor,
  B = d$block,
  T = d$treatment
)

m13.4 <- ulam(
  alist(
    P ~ bernoulli(p),
    logit(p) <- a[A] + b[T,B],
    # adaptive priors
    a[A] ~ dnorm(a_bar, sigma_a),
    matrix[T,B]:b ~ dnorm(0, sigma_b),
    # hyperpriors
    a_bar ~ dnorm(0, 1.5),
    c(sigma_a, sigma_b) ~ dexp(1)
  ), data = dat, chains = 4, cores = 4
)

precis(m13.4, 3, pars = c("a_bar", "sigma_a", "sigma_b"))
```

Plot the posterior distribution

```{r}
post <- extract.samples(m13.4)

# actor (handedness) effect
mean_a <- apply(inv_logit(post$a), 2, mean)
PI_a <- apply(inv_logit(post$a), 2, PI)
plot(NULL, xlab = "actor", ylab = "probability pull left",
     xlim = c(1, 7), ylim = c(0,1))
abline(h = 0.5, lty = 2)
points(1:7, mean_a, lwd = 4, cex = 1, col = "orchid")
for(i in 1:length(mean_a)) lines(c(i,i), c(PI_a[1,i], PI_a[2,i]), lwd = 7, col = col.alpha("orchid", 0.5))

# variation
dens(post$sigma_a, lwd = 3, col = "orchid", xlab = "sigma A")
```

Four actors appeared to be right-handed, and three actors appeared to be left-handed.

```{r}
# treatment effects
mean_b <- apply(post$b, c(2,3), mean)
plot(NULL, xlab = "treatment", ylab = "log-odds",
     xlim = c(0.5, 4.5), ylim = c(-2,2), xaxt = "n")
axis(1, at = c(1, 2, 3, 4), labels = c("R/N", "L/N", "R/P", "L/P"))
abline(h = 0, lty = 2)
for(i in 1:4) {
  for (j in 1:6) points(i+runif(1, -.15,.15), mean_b[i,j], cex = 1, lwd = 3, col = "seagreen")
}

# variation
dens(post$sigma_b, lwd = 3, col = "seagreen", xlab = "sigma B")
```


Treatment/block has little effects and less variation than the actor effect.


## Posterior prediciton of mutilevel models

When doing posterior predictions for multilevel models, we need to carefully consider how to incorporate the varying effects into the simulation, especially when using a hypothetical new population. We don't want to use the structure in the original data; rather we will simulate the varying effects from a distribution of the population.

```{r, include=FALSE}
data(reedfrogs)
d <- reedfrogs

dat <- list(
    S = d$surv,
    D = d$density,
    T = 1:nrow(d),
    P = ifelse( d$pred=="no" , 1L , 2L ),
    G = ifelse( d$size=="small" , 1L , 2L )
)
```

We use the reedfrog example. We will try to explore the causal effect of changing size distribution: Target population high density 35, 50% groups have predation. 25% large. Intervention makes 75% large. What is the change in survival?

```{r}
mSPG <- ulam(
  alist(
      S ~ binomial(D ,p),
      logit(p) <- a[T] + b[P,G],
      a[T] ~ normal(0, sigma),
      matrix[P,G]:b ~ normal(0, 1),
      sigma ~ exponential(1)
  ), data=dat, chains=4, cores=4)

post <- extract.samples(mSPG)

# sim first under status quo
n_groups <- 1000
n_samples <- 2000
S1 <- matrix(0, nrow = n_samples, ncol = n_groups)
for ( s in 1:n_groups ) {
    # sim a tank from posterior population
    aT <- rnorm(n_samples,0,post$sigma)
    # sample P and G for this group
    P <- sample(1:2, size=1, prob = c(0.5, 0.5)) # 50% pred
    G <- sample(1:2, size=1, prob = c(0.75, 0.25)) # 25% large
    # sim survival
    p <- inv_logit(aT + post$b[,P,G])
    S1[,s] <- rbinom(n_samples, 35, p)
}

# intervention - 50% large
S2 <- matrix(0, nrow = n_samples, ncol = n_groups)
for ( s in 1:n_groups ) {
    # sim a tank from posterior population
    aT <- rnorm(n_samples, 0, post$sigma)
    # sample P and G for this group
    P <- sample(1:2 ,size=1, prob=c(0.5,0.5)) # 50% pred
    G <- sample(1:2 ,size=1, prob=c(0.25,0.75)) # 75% large
    # sim survival
    p <- inv_logit(aT + post$b[,P,G])
    S2[,s] <- rbinom(n_samples, 35, p)
}

simplehist(as.numeric(S1), lwd=4, col=2, xlab="number surviving")
simplehist(as.numeric(S2), lwd=4, col=2, xlab="number surviving")

x <- table(as.numeric(S2 - S1))
x <- x/sum(x)

plot(NULL, xlim = c(-35,35), ylim = c(0,max(x)), 
     xlab = "change in survival", ylab = "proportion of simulation")
for (i in 1:length(x)) {
    xvals <- as.numeric(names(x))
    lines(rep(xvals[i], 2), c(0, x[i]), lwd = 4, col = ifelse(xvals[i]<0, 2, 4))
}

abline(v = mean(as.numeric(S2 - S1)), lty = 3, lwd = 3)

```


## Sampling issues in multilevel models

Multilevel models often have joint distributions with steep surfaces, which can make HMC sample inefficiently. The solution is to use non-centered parameters instead of centered parameters.

Centered parameters:

$v \sim \mathrm{Normal}(0, 3)$

$x \sim \mathrm{Normal}(0, \mathrm{exp}(v))$

Non-centered parameters:

$v \sim \mathrm{Normal}(0, 3)$

$z \sim \mathrm{Normal}(0, 1)$

$x = z\mathrm{exp}(v)$

The two are equivalent. The non-centered parameters essentially ask HMC to sample from a joint distribution of two Gaussian distributions and construct the variable afterwards. This makes sampling more efficient.

In the context of the chimp model, we will write this:

$P_i \sim \mathrm{Bernoulli}(p_i)$

$$\begin{aligned}
\mathrm{logit}(p_i) &= \bar{\alpha} + z_{\alpha, A[i]}\sigma_{\alpha} \\
&+ z_{\beta, T[i], B[i]}\sigma_{\beta}
\end{aligned}$$

$z_{\alpha, j} \sim \mathrm{Normal}(0, 1)$

$z_{\beta, j} \sim \mathrm{Normal}(0, 1)$

$\bar{\alpha} \sim \mathrm{Normal}(0, 1.5)$

$\sigma_{\alpha}, \sigma_{\beta} \sim \mathrm{Exponential}(1)$

```{r, include=FALSE}
d <- chimpanzees
d$treatment <- 1 + d$prosoc_left + 2*d$condition
dat <- list(
  P = d$pulled_left,
  A = d$actor,
  B = d$block,
  T = d$treatment
)
```


```{r}
m13.4nc <- ulam(
  alist(
    P ~ bernoulli(p),
    logit(p) <- a_bar + z_a[A]*sigma_a + z_b[T,B]*sigma_b,
    # adaptive priors
    z_a[A] ~ dnorm(0, 1),
    matrix[T,B]:z_b ~ dnorm(0, 1),
    # hyperpriors
    a_bar ~ dnorm(0, 1.5),
    c(sigma_a, sigma_b) ~ dexp(1),
    # generated quantities
    gq> vector[A]:a <<- a_bar + z_a*sigma_a,
    gq> matrix[T,B]:b <<- z_b*sigma_b
  ), data = dat, chains = 4, cores = 4
)
```

Plot the effective sample size against the centered model.

```{r}
the_pars <- c("a_bar", "sigma_a", "sigma_b", "b", "a")
precis_c <- precis(m13.4, depth = 3, pars = the_pars)
precis_nc <- precis(m13.4nc, depth =3 , pars = the_pars)
neff_table <- cbind(precis_c[,"n_eff"], precis_nc[,"n_eff"])
cols <- ifelse(neff_table[,1] > neff_table[,2], 2, 4)
plot(neff_table, xlim = range(neff_table), ylim = range(neff_table),
    xlab = "effective sample size (centered)", ylab = "effective sample size (non-centered)",
    lwd = 3, col = cols)
abline(a = 0, b = 1, lty = 2)

```

