---
title: 'Lecture 15: Social Networks'
author: "Yurun (Ellen) Ying"
date: '2022-06-29'
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(rethinking); library(dagitty); library(igraph);library(ellipse)
```

## Modeling Social Networks

The data we are gonna look at is a food sharing data in Arang Dak. The estimand is to what extent sharing is explained by reciprocity, aka, to what extent sharing is balanced within dyads. And how much sharing is explained by generalized sharing?

```{r}
data("KosterLeckie")
```

A DAG for this data (and perhaps for a lot of social network problems as well). Giving is motivated by the social ties in both directions as well as by general characteristics of the two households. The social ties are also influenced by general characteristics of the two households. These characteristics include things like wealth, hobbies, locations, etc.

```{r}
dag1 <- dagitty("dag{H_A -> G_AB <- H_B; T_AB -> G_AB <- T_BA;
                H_A -> T_AB <- H_B; H_A -> T_BA <- H_B}")
coordinates(dag1) <- list(x = c(H_A = 0.4, H_B = 0.6, G_AB = 0.5, T_AB = 0.5, T_BA = 0.5),
                          y = c(H_A = 0.5, H_B = 0.5, G_AB = 0.5, T_AB = 0.3, T_BA = 0.7))
drawdag(dag1)
```

### A generative model of social networks

We will now try to simulate social network data, build statistical models, and test them against the simulated data before we apply them on the real data. This process will be iterated many times in the workflow of real projects.

```{r}
# N households
N <- 25
dyads <- t(combn(N,2)) # all possible combinations of households
n_dyads <- nrow(dyads)

# simulate friendships in which ties are reciprocal
f <- rbern(n_dyads, 0.1) # 10% of dyads are friends

# simulate directed ties that are not reciprocal
alpha <- -3 # base rate of ties: -3 = 0.05
y <- matrix(NA, N, N) # matrix of ties for all individual households

# simulate ties
for (i in 1:N) for (j in 1:N) {
  
  if (i != j) {
  
    # directed tie from i to j
    ids <- sort(c(i,j))
    the_dyad <- which(dyads[,1] == ids[1] & dyads[,2] == ids[2])
    # either friend, or not friend but have a tie at a prob of 5%
    p_tie <- f[the_dyad] + (1 - f[the_dyad])*inv_logit(alpha) 
    y[i,j] <- rbern(1, p_tie)
  }
  
}

# simulate gifts
giftsAB <- rep(0, n_dyads)
giftsBA <- rep(0, n_dyads)
lambda <- log(c(0.5,2)) # rates of giving for y=0, y=1
for (i in 1:n_dyads){
  A <- dyads[i,1]
  B <- dyads[i,2]
  giftsAB[i] <- rpois(1, exp(lambda[1 + y[A,B]]))
  giftsBA[i] <- rpois(1, exp(lambda[1 + y[B,A]]))
}

# draw network
sng <- graph_from_adjacency_matrix(y)
lx <- layout_nicely(sng)
vcol <- "#DE536B"
plot(sng, layout = lx, vertex.size = 8, vertex.color = vcol,
     edge.arrow.size = 0.75, edge.width = 2, edge.curved = 0.35, edge.color = grau(),
     asp = 0.9, margin = -0.05, vertex.label = NA)
```

Now let's build a statistical model for this synthetic data. This time the model will not fully reflect what we have in the data. We will have to GLMs for gifts from A to B and those from B to A separately:

$$\begin{aligned}
G_{AB} &\sim \mathrm{Poisson}(\lambda_{AB}) \\
\mathrm{log}(\lambda_{AB}) &= \alpha + T_{AB} \\
G_{BA} &\sim \mathrm{Poisson}(\lambda_{BA}) \\
\mathrm{log}(\lambda_{BA}) &= \alpha + T_{BA} \\
\begin{pmatrix}
T_{AB} \\
T_{BA}
\end{pmatrix} &\sim \mathrm{MVNormal}(
\begin{bmatrix}
0 \\
0
\end{bmatrix},
\begin{bmatrix}
\sigma^2 & \rho \sigma^2 \\
\rho \sigma^2 & \sigma^2
\end{bmatrix}
)\\
\alpha &\sim \mathrm{Normal}(0,1) \\
\sigma &\sim \mathrm{Exponential}(1)\\
\rho &\sim \mathrm{LKJCorr(2)} \\
\end{aligned}$$

What we would like to know, really, is how much $T_{AB}$ and $T_{BA}$ are correlated.

```{r}
# dyad model
f_dyad <- alist(
  GAB ~ poisson(lambdaAB),
  GBA ~ poisson(lambdaBA),
  log(lambdaAB) <- a + T[D,1],
  log(lambdaBA) <- a + T[D,2],
  a ~ normal(0,1),
  
  # dyad effects
  transpars> matrix[n_dyads,2]:T <- compose_noncentered(rep_vector(sigma_T,2), L_Rho_T, Z),
  matrix[2,n_dyads]:Z ~ normal(0,1),
  cholesky_factor_corr[2]:L_Rho_T ~ lkj_corr_cholesky(2),
  sigma_T ~ exponential(1),
  # compute correlation matrix for dyads
  gq> matrix[2,2]:Rho_T <<- Chol_to_Corr(L_Rho_T)
)

# synthetic data
sim_data <- list(
    n_dyads = n_dyads,
    n_households = N,
    D = 1:n_dyads,
    HA = dyads[,1],
    HB = dyads[,2],
    GAB = giftsAB,
    GBA = giftsBA )
```

```{r, results='hide'}
# run the model with ulam
mGD <- ulam(f_dyad, data = sim_data, chains = 4, cores = 4, iter = 2000)
```

Let's do some plotting:

```{r}
post <- extract.samples(mGD)
# get the ties from A to B
tie_AB <- c()
tie_BA <- c()
for (i in 1:N-1) {
  tie_AB <- c(tie_AB, y[i,(i+1):25])
  tie_BA <- c(tie_BA, y[(i+1):25,i])
}
# posterior mean T
T_tie_mu <- apply(
  cbind(post$T[,which(tie_AB == 1),1], post$T[,which(tie_BA == 1),2]),
  2, mean)
T_notie_mu <- apply(
  cbind(post$T[,which(tie_AB == 0),1], post$T[,which(tie_BA == 0),2]),
  2, mean)

dens(T_tie_mu, col = "tomato", lwd = 3, ylim = c(0,5))
dens(T_notie_mu, col = "steelblue", lwd = 3, add = TRUE)
text(0.2, 2, "no tie", col = "steelblue")
text(0.6, 1, "tie", col = "tomato")
```

The posterior mean T for dyads without a tie is not different from 0, while the posterior mean T for dyads with a tie has most of its probability mass above 0.

```{r}
dens(post$Rho_T[,1,2], col = "olivedrab", lwd = 3, 
     xlim = c(-1,1), xlab = "correlation within dyads")
abline(v = 0, lty = 2)
```

The posterior distribution of the correlation within dyads has most of its probability mass above 0, indicating strong reciprocity of giving within dyads.

```{r}
# posterior mean T for AB and BA separately
TAB_mu <- apply(post$T[,,1], 2, mean)
TBA_mu <- apply(post$T[,,2], 2, mean)
plot(NULL, 
     xlim = range(TAB_mu)+c(-0.2,0.2), ylim = range(TBA_mu)+c(-0.2,0.2),
     xlab = "TAB", ylab = "TBA")
# color points if a tie is reciprocal
for (i in 1:length(TAB_mu))
  points(TAB_mu[i], TBA_mu[i], cex = 0.7, lwd = 2,
         col = ifelse(i %in% which(tie_AB==1 & tie_BA==1),
         "tomato", "steelblue"))
```

Plotting TAB and TBA helps us to see the positive association between the two. When a social tie is reciprocal, both households tend to give more.

### Apply to the sample

```{r}
# analyze sample
kl_data <- list(
    n_dyads = nrow(kl_dyads),
    n_households = max(kl_dyads$hidB),
    D = 1:nrow(kl_dyads),
    HA = kl_dyads$hidA,
    HB = kl_dyads$hidB,
    GAB = kl_dyads$giftsAB,
    GBA = kl_dyads$giftsBA )
```

```{r, results='hide'}
mGDkl <- ulam(f_dyad, data = kl_data, chains = 4, cores = 4, iter = 4000)
```

```{r}
precis(mGDkl, depth = 3, pars = c("a", "Rho_T", "sigma_T"))
```

We can see there is also a positive association within dyads in the real data.


### A generative model with general giving

What we have done so far, from the perspective of our DAG, is to model the effect of social network ties on giving. We haven't inspected the general giving in our data yet, which are the confounds. These come from the general features of households, such as wealth. As the previous section, we will do it first through a generative simulation.

```{r}
# simulate wealth
set.seed(238)
W <- rnorm(N) # standardized relative wealth in the community
bWG <- 0.5 # effect of wealth on giving - rich give more
bWR <- -1 # effect of wealth on receiving - rich receive less and poor get more

# simulate gifts
giftsAB <- rep(0, n_dyads)
giftsBA <- rep(0, n_dyads)
lambda <- log(c(0.5,2)) # rates of giving for y=0, y=1
for (i in 1:n_dyads){
  A <- dyads[i,1]
  B <- dyads[i,2]
  giftsAB[i] <- rpois(1, exp(lambda[1 + y[A,B]] + bWG*W[A] + bWR*W[B]))
  giftsBA[i] <- rpois(1, exp(lambda[1 + y[B,A]] + bWG*W[B] + bWR*W[A]))
}
```

We will also need to modify our model:

$$\begin{aligned}
G_{AB} &\sim \mathrm{Poisson}(\lambda_{AB}) \\
\mathrm{log}(\lambda_{AB}) &= \alpha + T_{AB} + G_A + R_B\\
G_{BA} &\sim \mathrm{Poisson}(\lambda_{BA}) \\
\mathrm{log}(\lambda_{BA}) &= \alpha + T_{BA} + G_B + R_A\\
\begin{pmatrix}
T_{AB} \\
T_{BA}
\end{pmatrix} &\sim \mathrm{MVNormal}(
\begin{bmatrix}
0 \\
0
\end{bmatrix},
\begin{bmatrix}
\sigma^2 & \rho \sigma^2 \\
\rho \sigma^2 & \sigma^2
\end{bmatrix}
)\\
\alpha &\sim \mathrm{Normal}(0,1) \\
\sigma &\sim \mathrm{Exponential}(1)\\
\rho &\sim \mathrm{LKJCorr(2)} \\
\begin{pmatrix}
G_A \\
R_A
\end{pmatrix} &\sim \mathrm{MVNormal}(
\begin{bmatrix}
0 \\
0
\end{bmatrix},
\begin{bmatrix}
\sigma_G^2 & r \sigma_G \sigma_R \\
r \sigma_G \sigma_R & \sigma_R^2
\end{bmatrix}
)\\
\sigma_G,\sigma_R &\sim \mathrm{Exponential}(1)\\
r &\sim \mathrm{LKJCorr(2)} \\
\end{aligned}$$

```{r}
# general model
f_general <- alist(
  GAB ~ poisson(lambdaAB),
  GBA ~ poisson(lambdaBA),
  log(lambdaAB) <- a + T[D,1] + gr[HA,1] + gr[HB,2],
  log(lambdaBA) <- a + T[D,2] + gr[HB,1] + gr[HA,2],
  a ~ normal(0,1),
  
  # dyad effects
  transpars> matrix[n_dyads,2]:T <- compose_noncentered(rep_vector(sigma_T,2), L_Rho_T, Z),
  matrix[2,n_dyads]:Z ~ normal(0,1),
  cholesky_factor_corr[2]:L_Rho_T ~ lkj_corr_cholesky(2),
  sigma_T ~ exponential(1),
  
  # general effects
  transpars> matrix[n_households,2]:gr <- compose_noncentered(sigma_gr, L_Rho_gr, Zgr),
  matrix[2,n_households]:Zgr ~ normal(0,1),
  cholesky_factor_corr[2]:L_Rho_gr ~ lkj_corr_cholesky(2),
  vector[2]:sigma_gr ~ exponential(1),
  
  # compute correlation matrix for dyads
  gq> matrix[2,2]:Rho_T <<- Chol_to_Corr(L_Rho_T),
  gq> matrix[2,2]:Rho_gr <<- Chol_to_Corr(L_Rho_gr)
)

# synthetic data
sim_data <- list(
    n_dyads = n_dyads,
    n_households = N,
    D = 1:n_dyads,
    HA = dyads[,1],
    HB = dyads[,2],
    GAB = giftsAB,
    GBA = giftsBA )
```

```{r, results='hide'}
# run the model with ulam
mGDR <- ulam(f_general, data = sim_data, chains = 4, cores = 4, iter = 2000)
```

```{r}
precis(mGDR, depth = 3, pars = c("a", "Rho_T", "sigma_T", "Rho_gr", "sigma_gr"))
```

Do some plotting to see the results

```{r}
# generalized giving against receiving
postGDR <- extract.samples(mGDR)
post_gg <- sapply(1:N, function(i) exp(postGDR$a + postGDR$gr[,i,1]))
gg_mu <- apply(post_gg, 2, mean)
post_gr <- sapply(1:N, function(i) exp(postGDR$a + postGDR$gr[,i,2]))
gr_mu <- apply(post_gr, 2, mean)

xymax <- max(gg_mu,gr_mu) + 0.5
plot(NULL, xlim = c(0,xymax), ylim = c(0,xymax),
     xlab = "generalized giving", ylab = "generalized receiving")
abline(a = 0, b = 1, lty = 2)
points(gg_mu, gr_mu, lwd = 3, cex = 0.8, col = "dodgerblue")
for (i in 1:N) {
  sigma <- cov(cbind(post_gg[,i], post_gr[,i]))
  mu <- c(gg_mu[i], gr_mu[i])
  lines(
    ellipse(sigma, centre = mu, level = 0.5), 
    col = col.alpha("black", 0.2))
}

# correlation between giving and receiving
dens(postGDR$Rho_gr[,1,2], xlim = c(-1,1), xlab = "correlation giving-receiving",
     lwd = 3, col = "cornflowerblue")
```


These two plots show that generalized giving and generalized receiving are negatively correlated. Those who give more tend to receive less.


```{r}
# correlation social tie
dens(postGDR$Rho_T[,1,2], xlim = c(-1,1), xlab = "correlation within dyads",
     lwd = 3, col = "olivedrab")
```

The correlation within dyads is even stronger in this model, being pushed all the way near 1.


### The real data with generalized giving and receiving

```{r, results='hide'}
mGDRkl <- ulam(f_general, data = kl_data, chains = 4, cores = 4, iter = 4000)
```

```{r}
# generalized giving against receiving
postGDRkl <- extract.samples(mGDRkl)
post_gg <- sapply(1:N, function(i) exp(postGDRkl$a + postGDRkl$gr[,i,1]))
gg_mu <- apply(post_gg, 2, mean)
post_gr <- sapply(1:N, function(i) exp(postGDRkl$a + postGDRkl$gr[,i,2]))
gr_mu <- apply(post_gr, 2, mean)

xymax <- max(gg_mu,gr_mu) + 0.5
plot(NULL, xlim = c(0,xymax), ylim = c(0,xymax),
     xlab = "generalized giving", ylab = "generalized receiving")
abline(a = 0, b = 1, lty = 2)
points(gg_mu, gr_mu, lwd = 3, cex = 0.8, col = "dodgerblue")
for (i in 1:N) {
  sigma <- cov(cbind(post_gg[,i], post_gr[,i]))
  mu <- c(gg_mu[i], gr_mu[i])
  lines(
    ellipse(sigma, centre = mu, level = 0.5), 
    col = col.alpha("black", 0.2))
}

# correlation between giving and receiving
dens(postGDRkl$Rho_gr[,1,2], xlim = c(-1,1), xlab = "correlation giving-receiving",
     lwd = 3, col = "cornflowerblue")

# correlation social tie
dens(postGDRkl$Rho_T[,1,2], xlim = c(-1,1), xlab = "correlation within dyads",
     lwd = 3, col = "olivedrab")
```

Same pattern here: generalized giving and receiving are negative correlated, and correlation within dyads is strong.


### Adding linear model for social tie, giving, and receiving

The previous models treat the social ties, generalized giving and receiving as single varying effects. They are really just placeholders for other causes, so we can also model them as linear functions of the causes we are interested in. In our case, this is the strength of social ties and wealth for giving and receiving.

Let's turn our mmodel into some fancy stuff:

$$\begin{aligned}
G_{AB} &\sim \mathrm{Poisson}(\lambda_{AB}) \\
\mathrm{log}(\lambda_{AB}) &= \alpha + \mathcal{T}_{AB} + \mathcal{G}_A + \mathcal{R}_B\\
\mathcal{T}_{AB} &= T_AB + \beta_{A} A_{AB}\\
\mathcal{G}_A &= G_A + \beta_{W,G} W_{A}\\
\mathcal{R}_B &= R_B + \beta_{W,R} W_{B}\\
G_{BA} &\sim \mathrm{Poisson}(\lambda_{BA}) \\
\mathrm{log}(\lambda_{BA}) &= \alpha + \mathcal{T}_{BA} + \mathcal{G}_B + \mathcal{R}_A\\
\mathcal{T}_{BA} &= T_BA + \beta_{A} A_{BA}\\
\mathcal{G}_B &= G_B + \beta_{W,G} W_{B}\\
\mathcal{R}_A &= R_A + \beta_{W,R} W_{A}\\
\end{aligned}$$

```{r}
# model with household features
f_house <- alist(
  # A toB
  GAB ~ poisson(lambdaAB),
  log(lambdaAB) <- a + TAB + GA + RB,
  TAB <- T[D,1] + bA*A,
  GA <- gr[HA,1] + bW[1]*W[HA],
  RB <- gr[HB,2] + bW[2]*W[HB],
  
  GBA ~ poisson(lambdaBA),
  log(lambdaBA) <- a + TBA + GB + RA,
  TBA <- T[D,2] + bA*A,
  GB <- gr[HB,1] + bW[1]*W[HB],
  RA <- gr[HA,2] + bW[2]*W[HA],
  
  # priors
  a ~ normal(0, 1),
  bA ~ normal(0, 1),
  vector[2]:bW ~ normal(0,1),
  
  # dyad effects
  transpars> matrix[n_dyads,2]:T <- compose_noncentered(rep_vector(sigma_T,2), L_Rho_T, Z),
  matrix[2,n_dyads]:Z ~ normal(0,1),
  cholesky_factor_corr[2]:L_Rho_T ~ lkj_corr_cholesky(2),
  sigma_T ~ exponential(1),
  
  # general effects
  transpars> matrix[n_households,2]:gr <- compose_noncentered(sigma_gr, L_Rho_gr, Zgr),
  matrix[2,n_households]:Zgr ~ normal(0,1),
  cholesky_factor_corr[2]:L_Rho_gr ~ lkj_corr_cholesky(2),
  vector[2]:sigma_gr ~ exponential(1),
  
  # compute correlation matrix for dyads
  gq> matrix[2,2]:Rho_T <<- Chol_to_Corr(L_Rho_T),
  gq> matrix[2,2]:Rho_gr <<- Chol_to_Corr(L_Rho_gr)
)


# compose data
kl_data <- list(
    n_dyads = nrow(kl_dyads),
    n_households = max(kl_dyads$hidB),
    D = 1:nrow(kl_dyads),
    HA = kl_dyads$hidA,
    HB = kl_dyads$hidB,
    GAB = kl_dyads$giftsAB,
    GBA = kl_dyads$giftsBA,
    W = standardize(kl_households$hwealth), # wealth
    A = standardize(kl_dyads$dass) ) # association index - time spent together
```

```{r, results='hide'}
# fit the model
mGDRAkl <- ulam(f_house, data = kl_data, chains = 4, cores = 4, iter = 4000)
```

```{r}
postGDRAkl <- extract.samples(mGDRAkl)
# the effect of wealth
plot(NULL, xlim = c(-0.5, 0.8), ylim = c(0,4.5), xlab = "Effect of wealth")
dens(postGDRAkl$bW[,1], lwd = 3, col = "dodgerblue", add = TRUE)
dens(postGDRAkl$bW[,2], lwd = 3, col = "tomato", add = TRUE)
abline(v = 0, lty = 2)
text(0.5, 1, "giving", col = "dodgerblue")
text(-0.4, 3, "receiving", col = "tomato")
```

Th effect of wealth on giving is larger positive, while its effect on receiving is negative. The rich gives more and receive less.