---
title: 'Lecture 7.5: Interaction'
author: "Yurun (Ellen) Ying"
date: "6/5/2022"
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rethinking)
```


## Interaction between a cateogorical and a continuous variable

Let's investigate how ruggedness of a country is associated with its GDP, and how this association depends on continents.

```{r}
data(rugged)
d <- rugged

# make a log version of GDP
# GDP grows exponential - more wealth generates more wealth
# logged GDP is more likely to be linearly associated with other variables
d$log_gdp <- log(d$rgdppc_2000)

# extract countries with gdp datd
d2 <- d[complete.cases(d$rgdppc_2000),]

# standardize the variables
# gdp standardized so that 1 represents global average
d2$log_gdp_std <- d2$log_gdp / mean(d2$log_gdp)
# ruggedness standardized so that the scale is from 0 to 1
d2$rugged_std <- d2$rugged/max(d2$rugged)

m_rugged <- mean(d2$rugged_std)
```

### First model: a linear model of logged GDP regressed on ruggedness.

```{r}
m8.1 <- quap(
  alist(
    log_gdp_std ~ dnorm(mu, sigma),
    mu <- a + b * (rugged_std - m_rugged),
    a ~ dnorm(1, 1),
    b ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = d2
)

# let's investigate the priors first
set.seed(7)
prior <- extract.prior(m8.1)

plot(NULL, xlim = c(0, 1), ylim = c(0.5, 1.5),
     xlab = "ruggedness", ylab = "logged GDP")
abline(h = min(d2$log_gdp_std), lty = 2)
abline(h = max(d2$log_gdp_std), lty = 2)

# draw 50 lines for the prior
r_seq <- seq(-0.1, 1.1, length.out = 30)
mu <- link(m8.1, post = prior, data = list(rugged_std = r_seq))
for(i in 1:50) lines(r_seq, mu[i,], col = col.alpha("black", 0.3))
```

Note that this set of priors doesn't make much sense since many prior predictions lie outside of possible values. Let's use more skeptcial priors.

```{r}
m8.1 <- quap(
  alist(
    log_gdp_std ~ dnorm(mu, sigma),
    mu <- a + b * (rugged_std - m_rugged),
    a ~ dnorm(1, 0.1),
    b ~ dnorm(0, 0.3),
    sigma ~ dexp(1)
  ),
  data = d2
)
```

And we can see these priors give lines that are not too extreme.

```{r, echo=FALSE}
# let's investigate the priors first
set.seed(7)
prior <- extract.prior(m8.1)

plot(NULL, xlim = c(0, 1), ylim = c(0.5, 1.5),
     xlab = "ruggedness", ylab = "logged GDP")
abline(h = min(d2$log_gdp_std), lty = 2)
abline(h = max(d2$log_gdp_std), lty = 2)

# draw 50 lines for the prior
r_seq <- seq(-0.1, 1.1, length.out = 30)
mu <- link(m8.1, post = prior, data = list(rugged_std = r_seq))
for(i in 1:50) lines(r_seq, mu[i,], col = col.alpha("black", 0.3))
```

```{r}
precis(m8.1)
```

### Second model: Include being in Africa or not as a predictor

```{r}
# make variable to index Africa (1) or not (2)
d2$cid <- ifelse(d2$cont_africa == 1, 1, 2)

m8.2 <- quap(
  alist(
    log_gdp_std ~ dnorm(mu, sigma),
    mu <- a[cid] + b * (rugged_std - m_rugged),
    a[cid] ~ dnorm(1, 0.1),
    b ~ dnorm(0, 0.3),
    sigma ~ dexp(1)
  ),
  data = d2
)

# compare models
# the model including Africa does get better in prediction
compare(m8.1, m8.2)
```

A closer look at what's happening in this model

```{r}
precis(m8.2, depth = 2)

# contrast and its confidence interval
post <- extract.samples(m8.2)
diff_a1_a2 <- post$a[,1] - post$a[,2] 
mean(diff_a1_a2); PI(diff_a1_a2)
```

This means that the difference in the mean GDP between African and non-African countries is reliably below zero.

```{r}
mu_Africa <- link(m8.2, data = data.frame(cid = 1, rugged_std = r_seq))
mu_Africa_mu <- apply(mu_Africa, 2, mean)
mu_Africa_PI <- apply(mu_Africa, 2, PI, prob = .97)
mu_nonAfrica <- link(m8.2, data = data.frame(cid = 2, rugged_std = r_seq))
mu_nonAfrica_mu <- apply(mu_nonAfrica, 2, mean)
mu_nonAfrica_PI <- apply(mu_nonAfrica, 2, PI, prob = .97)

# plot the predictions against data
plot(NULL, xlim = c(0, 1), ylim = c(0.5, 1.5),
     xlab = "ruggedness", ylab = "logged GDP")
points(d2$rugged_std[d2$cid == 1], d2$log_gdp_std[d2$cid == 1],
       col = rangi2, pch = 16)
lines(r_seq, mu_Africa_mu, col = rangi2, lwd = 3)
shade(mu_Africa_PI, r_seq, col = col.alpha(rangi2, 0.2))
points(d2$rugged_std[d2$cid == 2], d2$log_gdp_std[d2$cid == 2],
       col = "gray", pch = 16)
lines(r_seq, mu_nonAfrica_mu, lwd = 3)
shade(mu_nonAfrica_PI, r_seq)
```

The mean GDP outside of Africa is higher than the mean GPD in Africa, but the slopes are roughly the same. This means little.

### Third model: Slope also depends on being in Africa or not.

```{r}
m8.3 <- quap(
  alist(
    log_gdp_std ~ dnorm(mu, sigma),
    mu <- a[cid] + b[cid] * (rugged_std - m_rugged),
    a[cid] ~ dnorm(1, 0.1),
    b[cid] ~ dnorm(0, 0.3),
    sigma ~ dexp(1)
  ),
  data = d2
)

# compare models
compare(m8.1, m8.2, m8.3, func = PSIS)
```

m8.3 has a weight larger than 0.95 and this is a strong support for choosing m8.3 if our goal is to make prediction. But the 0.03 weight given to m8.2 suggests that there is a little overfit in m8.3. And the difference between m8.2 and m8.3 is barely over 0. Maybe need to try robust regression at this point.

```{r, warning=FALSE}
# see the influential cases
plot(PSIS(m8.3, pointwise = TRUE)$k, ylim = c(-0.5, 1))
```

Now plot the interaction into separate plots.

```{r}
# African countries
mu_Africa <- link(m8.3, data = data.frame(cid = 1, rugged_std = r_seq))
mu_Africa_mu <- apply(mu_Africa, 2, mean)
mu_Africa_PI <- apply(mu_Africa, 2, PI, prob = .97)

plot(NULL, xlim = c(0, 1), ylim = c(0.5, 1.5),
     xlab = "ruggedness", ylab = "logged GDP")
points(d2$rugged_std[d2$cid == 1], d2$log_gdp_std[d2$cid == 1],
       col = rangi2, pch = 16)
lines(r_seq, mu_Africa_mu, col = rangi2, lwd = 3)
shade(mu_Africa_PI, r_seq, col = col.alpha(rangi2, 0.2))
mtext("African countries")

# non-African countries
mu_nonAfrica <- link(m8.3, data = data.frame(cid = 2, rugged_std = r_seq))
mu_nonAfrica_mu <- apply(mu_nonAfrica, 2, mean)
mu_nonAfrica_PI <- apply(mu_nonAfrica, 2, PI, prob = .97)

plot(NULL, xlim = c(0, 1), ylim = c(0.5, 1.5),
     xlab = "ruggedness", ylab = "logged GDP")
points(d2$rugged_std[d2$cid == 2], d2$log_gdp_std[d2$cid == 2],
       col = "gray", pch = 16)
lines(r_seq, mu_nonAfrica_mu, lwd = 3)
shade(mu_nonAfrica_PI, r_seq)
mtext("Non-African countries")
```

### Symmetry of interaction

There are two mathematically equivalent interpretations of one interaction effect.

- The association between ruggedness and GDP depends on whether the country is in Africa.
- The association between being African countries and GDP depends upon ruggedness.

Let's take a look at the second interpretation

```{r}
delta <- mu_Africa - mu_nonAfrica
delta_mu <- apply(delta, 2, mean)
delta_PI <- apply(delta, 2, PI, prob = 0.97)

plot(NULL, xlim = c(0, 1), ylim = c(-0.5, 0.5),
     xlab = "ruggedness", ylab = "expected difference in log GDP")
abline(h = 0, lty = 2)
lines(r_seq, delta_mu)
shade(delta_PI, r_seq)
```

African countries have lower GDP than non-African countries only when ruggedness is low. When ruggedness is high, there is little difference in the GDP of African and non-African countries.


## Continuous interactions

Interactions between continuous variables are harder to interpret. Let's see an example of tulip growth affected by water and shade.

```{r}
data("tulips")
d <- tulips

# standardization
d$blooms_std <- d$blooms / max(d$blooms) # to preserve the meaningful 0 value
d$water_cent <- d$water - mean(d$water) # centering
d$shade_cent <- d$shade - mean(d$shade)

```

It is reasonable to think alpha has a mean value of 0.5 (which is the mean of blossom). And the range of blossom is 0 to 1, so we can pick a prior with a sd of 0.25, which gives a distribution with 95% of likely values lying between 0 and 1. For the two slopes, we don't assume any directions so we choose a mean of 0. The range of water and shades is -1 to 1, which is 2 units. This means range of the slope will be -0.5 to 0.5, which implies a prior with a sd of 0.25.

```{r}
# a model without interaction
m8.4 <- quap(
  alist(
    blooms_std ~ dnorm(mu, sigma),
    mu <- a + bw*water_cent + bs*shade_cent,
    a ~ dnorm(0.5, 0.25),
    bw ~ dnorm(0, 0.25),
    bs ~ dnorm(0, 0.25),
    sigma ~ dexp(1)
  ),
  data = d
)

# a model with interaction
# hard to give the interaction term a sensible prior
# let's just use the same as the main effect terms
m8.5 <- quap(
  alist(
    blooms_std ~ dnorm(mu, sigma),
    mu <- a + bw*water_cent + bs*shade_cent + bws*water_cent*shade_cent,
    a ~ dnorm(0.5, 0.25),
    bw ~ dnorm(0, 0.25),
    bs ~ dnorm(0, 0.25),
    bws ~ dnorm(0, 0.25),
    sigma ~ dexp(1)
  ),
  data = d
)
```

Interpret the results with triptych graphs

```{r, fig.width=9, fig.height=4}
# without interaction
par(mfrow = c(1, 3))
for (s in -1:1) {
  idx <- which(d$shade_cent == s)
  plot(d$water_cent[idx], d$blooms_std[idx], xlim = c(-1, 1), ylim = c(0, 1),
       xlab = "water", ylab = "blooms", pch = 16, col = rangi2)
  mu <- link(m8.4, data = data.frame(shade_cent = s, water_cent = -1:1))
  for (i in 1:20) lines(-1:1, mu[i,], col = col.alpha("black", 0.3))
}
```

The slopes don't change with different values of shade, but the lines do sink down, meaning that shade has negative influences on tulips bloom.

```{r, fig.width=9, fig.height=4}
# with interaction
par(mfrow = c(1, 3))
for (s in -1:1) {
  idx <- which(d$shade_cent == s)
  plot(d$water_cent[idx], d$blooms_std[idx], xlim = c(-1, 1), ylim = c(0, 1),
       xlab = "water", ylab = "blooms", pch = 16, col = rangi2)
  mu <- link(m8.5, data = data.frame(shade_cent = s, water_cent = -1:1))
  for (i in 1:20) lines(-1:1, mu[i,], col = col.alpha("black", 0.3))
}

```


The slopes change with different values of shade, meaning that when light is weak, the effect of water on blooms also gets weak.